Parallel Nonnegative CP Decomposition \\ of Dense Tensors

### 2018-03-26 latex/sc18/experiments.tex
added ksweep plot
finished talk slides

>>>>

\begin{figure}
\renewcommand{\datafile}{data/ksw_4D_HSI.dat}
\renewcommand{\numiterations}{10}
\begin{tikzpicture}
\begin{axis}[	
	ybar stacked,
	bar width=8pt,
	width=\columnwidth,
	height =.75\columnwidth,
	%width=9cm, height=3.85cm,
	ylabel={Time (s)}, 
	xlabel={Rank $k$},
	y label style={yshift=-.5cm},
	ymin=0,
	%symbolic x coords={D10,N10,,D20,N20,,D30,N30,,D40,N40,,D50,N50},
	symbolic x coords={D10,D20,D30,D40,D50},
	%xtick=data,
	xticklabels={,10,20,30,40,50},
	legend style={at={(0.5,1.3)},anchor=north},
	legend columns=-1,
]
	\setcolors
	\addplot table[x=alg-K, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=alg-K, y expr=(\thisrow{mttv}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=alg-K, y expr=((\thisrow{nnls}+\thisrow{gram})/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=alg-K, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=alg-K, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
	\legend{PM,mTTV,NLS,Factor Comm,Gram Comm};
\end{axis}
%\labels
\end{tikzpicture}
\caption{Per-iteration time breakdown of our implementation (using dimension trees) over various ranks for a time-lapse HSI dataset with dimensions $1344\times 1024\times 33 \times 9$ on 64 processors arranged in a $8\times8\times1\times1$ grid.}
\end{figure}

### 2018-03-26 latex/sc18/plots.tex
finished talk slides

\newcommand{\setcolors}{
\pgfplotsset{cycle list={
	red, fill=red \\ 
	blue, fill=blue \\ 
	green, fill=green \\ 
	red, pattern=crosshatch, pattern color=red \\
	blue, pattern=crosshatch, pattern color=blue \\
	green, pattern=crosshatch, pattern color=green \\
}};
}

% set options for grouped bar plot
\newcommand{\breakdownplotoptions}{
	ybar stacked,
	reverse legend,
	bar width=8pt,
	width=9cm, height=3.85cm,
	ylabel={Time (s)}, 
	%y label style={yshift=-.5cm},
	ymin=0,
	symbolic x coords={D1,N1,,D16,N16,,D81,N81,,D256},
	xtick=data,
	legend style={at={(0.5,1.3)},anchor=north},
	legend columns=-1,
	reverse legend
}

>>>>

\newcommand{\setcolors}{
\pgfplotsset{cycle list={
	red, fill=red \\ 
	blue, fill=blue \\ 
	green, fill=green \\ 
	red, pattern=crosshatch, pattern color=red \\
	green, pattern=crosshatch, pattern color=green \\
	blue, pattern=crosshatch, pattern color=blue \\
}};
}

% set options for grouped bar plot
\newcommand{\breakdownplotoptions}{
	ybar stacked,
	reverse legend,
	bar width=8pt,
	width=9cm, height=3.85cm,
	ylabel={Time (s)}, 
	y label style={yshift=-.5cm},
	ymin=0,
	symbolic x coords={D1,N1,D16,N16,D81,N81,D256,N256},
	xtick=data,
	legend style={at={(0.5,1.3)},anchor=north},
	legend columns=-1,
	reverse legend
}

### 2018-03-26 latex/slides/talk.tex
finished talk slides

\usepackage{setspace}
\usepackage{beamerthemesplit}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{times}
\usepackage[T1]{fontenc}
\usepackage{color}
\usepackage{colortbl}
%\usepackage{algorithm,algorithmic}
\usepackage{multimedia,xmpmulti}
\usepackage{multirow,multicol}
\usepackage[absolute,overlay]{textpos}
\usepackage{multimedia,xmpmulti}
\usepackage{tabularx}
\usepackage{algorithm,algpseudocode}

@@

\footnotesize
\begin{itemize}
	\item Basic algorithm is block coordinate descent, each block is factor matrix
	\item For each block, solve nonnegative linear least squares problem
\end{itemize}
\normalsize

@@

%\begin{algorithm}
%\caption{$(\CPl,\epsilon) = \text{NNCP}(\TA,k)$}
%\caption{$\CP = \text{NNCP}(\TA,k)$}
%\label{alg:nncp}
\begin{algorithmic}[1]
\Require $\TA$ is $I_1\times \cdots \times I_N$ tensor, $k$ is approximation rank
%\State \Comment{Compute NNCP approximation}
\While{not converged}
	\State \Comment{Perform outer iteration of BCD}
	\For{$n=1$ to $N$}
	\State \Comment{Compute new factor matrix in $n$th mode}
	\State \red{$\Mn{M}{n} = \text{MTTKRP}(\TA,\{\Mn{H}{i}\},n)$}
	\State $\Mn{S}{n} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$
	\State \blue{$\Mn{H}{n} = \text{NLS-Update}(\Mn{S}{n},\Mn{M}{n})$}
%	\State $\Mn[\hat]{H}{n} = \text{NLS-Update}(\Mn{S}{n},\Mn{M}{n})$
%	\State \Comment{Normalize columns}
%	\State $\V{\lambda} = \text{Col-Norms}(\Mn[\hat]{H}{n})$
%	\State $\Mn{H}{n} = \text{Col-Scale}(\Mn[\hat]{H}{n},\V{\lambda})$
%	\State \Comment{Organize data for later modes}
	\State $\Mn{G}{n} = \MnTra{H}{n}\Mn{H}{n}$
	\EndFor
%	\State \Comment{Compute relative error $\epsilon$ from mode-$N$ matrices}
%	\State $\beta = \langle \Mn{M}{N},\Mn[\hat]{H}{N} \rangle$
%	\State $\gamma = \V{\lambda}^\Tra (\Mn{S}{N} \Hada \Mn{G}{N}) \V{\lambda}$
%	\State $\epsilon = \sqrt{(\alpha-2\beta+\gamma)/\alpha}$ 
\EndWhile
%\Ensure $\|\TA - \dsquare{\V{\lambda}; \Mn{H}{1},\dots,\Mn{H}{N}}\| /\|\TA\| = \epsilon$
\Ensure $\TA \approx \CP$
\end{algorithmic}
%\end{algorithm}

@@

\begin{frame}

>>>>

\usepackage{setspace}
\usepackage{beamerthemesplit}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{times}
\usepackage[T1]{fontenc}
\usepackage{color}
\usepackage{colortbl}
%\usepackage{algorithm,algorithmic}
\usepackage{multimedia,xmpmulti}
\usepackage{multirow,multicol}
\usepackage[absolute,overlay]{textpos}
\usepackage{multimedia,xmpmulti}
\usepackage{tabularx}
\usepackage{algorithm,algpseudocode}
\usepackage{subfig}

\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{etoolbox}
\usetikzlibrary{patterns}

@@

\footnotesize
\begin{itemize}
	\item Basic algorithm is block coordinate descent, each block is factor matrix
	\item For each block, solve nonnegative linear least squares problem
	\begin{itemize}
		\item \scriptsize assuming least squares loss function (Gaussian distribution)
	\end{itemize}
\end{itemize}
\normalsize

@@

\only<1>{
%\begin{algorithm}
%\caption{$(\CPl,\epsilon) = \text{NNCP}(\TA,k)$}
%\caption{$\CP = \text{NNCP}(\TA,k)$}
%\label{alg:nncp}
\begin{algorithmic}[1]
\Require $\TA$ is $I_1\times \cdots \times I_N$ tensor, $k$ is approximation rank
%\State \Comment{Compute NNCP approximation}
\While{not converged}
	\State \Comment{Perform outer iteration of BCD}
	\For{$n=1$ to $N$}
	\State \Comment{Compute new factor matrix in $n$th mode}
	\State \red{$\Mn{M}{n} = \text{MTTKRP}(\TA,\{\Mn{H}{i}\},n)$}
	\State $\Mn{S}{n} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$
	\State \blue{$\Mn{H}{n} = \text{NLS-Update}(\Mn{S}{n},\Mn{M}{n})$}
%	\State $\Mn[\hat]{H}{n} = \text{NLS-Update}(\Mn{S}{n},\Mn{M}{n})$
%	\State \Comment{Normalize columns}
%	\State $\V{\lambda} = \text{Col-Norms}(\Mn[\hat]{H}{n})$
%	\State $\Mn{H}{n} = \text{Col-Scale}(\Mn[\hat]{H}{n},\V{\lambda})$
%	\State \Comment{Organize data for later modes}
	\State $\Mn{G}{n} = \MnTra{H}{n}\Mn{H}{n}$
	\EndFor
%	\State \Comment{Compute relative error $\epsilon$ from mode-$N$ matrices}
%	\State $\beta = \langle \Mn{M}{N},\Mn[\hat]{H}{N} \rangle$
%	\State $\gamma = \V{\lambda}^\Tra (\Mn{S}{N} \Hada \Mn{G}{N}) \V{\lambda}$
%	\State $\epsilon = \sqrt{(\alpha-2\beta+\gamma)/\alpha}$ 
\EndWhile
%\Ensure $\|\TA - \dsquare{\V{\lambda}; \Mn{H}{1},\dots,\Mn{H}{N}}\| /\|\TA\| = \epsilon$
\Ensure $\TA \approx \CP$
\end{algorithmic}
%\end{algorithm}
}
\only<2>{
\begin{itemize}
	\item Lots of ways to solve nonnegative linear least squares 
	\begin{enumerate}
		\item multiplicative updates (one element at a time, Lee-Seung)
		\item hierarchical alternating least squares (one column at a time, Cichocki et al.)
		\item block principal pivoting (active set method, Park et al.)
		\item Nesterov algorithm (used by Sidiropolous et al.)
		\item alternating directions method of multipliers
	\end{enumerate}
	\item All are bottlenecked by MTTKRP and also need Gram 
	\begin{itemize}	
		\item MTTKRP bottleneck in dense case, maybe not sparse?
	\end{itemize}
\end{itemize}
}

@@

\begin{frame}
\frametitle{Parallel Algorithm for NNCP via BCD/ALS}

\footnotesize
\textbf{Data distribution}
\begin{itemize}
	\item block distribution of tensor over $N$-D logical processor grid
	\item row-wise distribution of each factor matrix 
\end{itemize}

\textbf{Communication steps}
\begin{itemize}
	\item All-Gather and Reduce-Scatter across processor slices for MTTKRP
	\item All-Reduce across all processors for Gram
	\item All-Reduce across all processors for normalization and error comp.
\end{itemize}

\textbf{Local Computation}
\begin{itemize}
	\item local MTTKRP (which we do using dimension trees)
	\item local NLS solves/updates (which we do using block principal pivoting)
	\item normalization and error computation (very cheap)
\end{itemize}

\input{../sc18/fig/Alg_header.tex}

\begin{frame}
\frametitle{Inner Iteration of Parallel Algorithm}

\begin{center}
\scalebox{1.5}{
\only<1>{\input{../sc18/fig/AlgStart.tex}}
\only<2>{\input{../sc18/fig/AlgMTTKRP.tex}}
\only<3>{\input{../sc18/fig/AlgRS.tex}}
\only<4>{\input{../sc18/fig/AlgNLS.tex}}
\only<5>{\input{../sc18/fig/AlgAG.tex}}
}

\only<1>{Start $n$th iteration with redundant subset of rows \\ of each input matrix.}
\only<2>{Compute local MTTKRP for contribution to output matrix $\Mn{M}{2}$.}
\only<3>{Reduce-Scatter to compute and distribute rows of $\Mn{M}{2}$.}
\only<4>{Compute local NLS update to obtain $\Mn{H}{2}_{\V{p}}$ from $\Mn{M}{2}_{\V{p}}$ (and $\Mn{S}{2}$).}
\only<5>{All-Gather to collect rows of $\Mn{H}{2}$ needed \\ for later inner iterations.}

\end{center}

\begin{frame}
\frametitle{Inner Iteration of Parallel Algorithm}

\begin{figure}
\centering
  \subfloat[Start]{\input{../sc18/fig/AlgStart.tex}} \quad
  \subfloat[Local MTTKRP]{\input{../sc18/fig/AlgMTTKRP.tex}} \quad
  \subfloat[Reduce-Scatter]{\input{../sc18/fig/AlgRS.tex}} \quad
  \subfloat[Local NLS]{\input{../sc18/fig/AlgNLS.tex}} \quad
  \subfloat[All-Gather]{\input{../sc18/fig/AlgAG.tex}}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Dimension Trees}

\begin{center}
\footnotesize
Dimension trees systematically avoid recomputation across MTTKRPs 
\normalsize

\input{../sc18/fig/dim_tree.tex}
\end{center}

\footnotesize
Nearly all the time spent in partial MTTKRP (PM), same number of flops as MTTKRP but easier to implement/parallelize (one DGEMM call)
\normalsize

\begin{frame}
\frametitle{Error Computation}

Given data tensor $\TA$ and CP model $\T{M} = \CP$, we want to compute the squared error
$$\| \TA - \T{M}\|^2 = \|\TA\|^2 - 2\langle \TA, \T{M} \rangle + \|\T{M}\|^2$$
which we do efficiently using the identity
$$\langle \TA, \T{M} \rangle = \langle \Mn{M}{N}, \Mn{H}{N} \rangle$$ 
where 
$$\Mn{M}{N} = \Mz{A}{N} (\Mn{H}{N-1} \Khat \cdots \Khat \Mn{H}{1})$$

*This technique is noted in a paper by Smith and Karypis and used in code by Phan, Tichavsky, and Cichocki

\input{../sc18/plots}

\begin{frame}
\frametitle{Strong Scaling on 3D Tensor}

\begin{figure}
\begin{tikzpicture}[scale=.9]
\renewcommand{\datafile}{../sc18/data/str_3D_syn.dat}
\renewcommand{\numiterations}{42}
\liavastrue
\strongscalingplot
\end{tikzpicture}
\caption{Strong scaling of 3D synthetic tensor with dimension $1024\times 1024\times 1024$ on processor grids $2^k\times 2^k\times 2^k$ for $k\in\{0,1,2,3\}$.  The rank is fixed at 32.}
\end{figure}

\begin{frame}
\frametitle{Strong Scaling on 5D Tensor}

\begin{figure}
\begin{tikzpicture}[scale=.9]
\renewcommand{\datafile}{../sc18/data/str_5D_syn.dat}
\renewcommand{\numiterations}{10}
\liavasfalse
\strongscalingplot
\end{tikzpicture}
\caption{Strong scaling of 5D synthetic tensor with dimension $64\times 64\times 64\times 64\times 64$ on processor grids $1\times1\times1\times1\times1$, $2\times1\times1\times1\times1$, $\dots$, $2\times2\times2\times2\times2$.  The rank is fixed at 32.}
\end{figure}

\begin{frame}
\frametitle{Weak Scaling Time Breakdown}

\begin{figure}
\begin{tikzpicture}
\renewcommand{\datafile}{../sc18/data/wk_4D_syn.dat}
\renewcommand{\numiterations}{10}
\breakdownplot
%\labels
\end{tikzpicture}
\caption{Weak scaling of 4D synthetic tensors with (D) and without (N) the use of dimension trees.  The tensor and processor grid dimensions are $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$, and the rank is fixed at 32.}
\end{figure}

\begin{frame}
\frametitle{Varying Rank Time Breakdown}

\begin{figure}
\renewcommand{\datafile}{../sc18/data/ksw_4D_HSI.dat}
\renewcommand{\numiterations}{10}
\begin{tikzpicture}[scale=.8]
\begin{axis}[	
	ybar stacked,
	bar width=12pt,
	width=\columnwidth,
	height =.5\columnwidth,
	%width=9cm, height=3.85cm,
	ylabel={Time (s)}, 
	xlabel={Rank $k$},
	y label style={yshift=-.5cm},
	ymin=0,
	%symbolic x coords={D10,N10,,D20,N20,,D30,N30,,D40,N40,,D50,N50},
	symbolic x coords={D10,D20,D30,D40,D50},
	%xtick=data,
	xticklabels={,10,,20,,30,,40,,50},
	legend style={at={(0.5,1.3)},anchor=north},
	legend columns=-1,
]
	\setcolors
	\addplot table[x=alg-K, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=alg-K, y expr=(\thisrow{mttv}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=alg-K, y expr=((\thisrow{nnls}+\thisrow{gram})/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=alg-K, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=alg-K, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
	\legend{PM,mTTV,NLS,Factor Comm,Gram Comm};
\end{axis}
%\labels
\end{tikzpicture}
\caption{Per-iteration time breakdown of our implementation (using dimension trees) over various ranks for a time-lapse HSI dataset with dimensions $1344\times 1024\times 33 \times 9$ on 64 processors arranged in a $8\times8\times1\times1$ grid.}
\end{figure}

\begin{frame}
\frametitle{Conclusions (so far)}

\footnotesize
\begin{enumerate}
	\item Hard not to be compute bound in dense case
	\begin{itemize}
	\scriptsize
		\item computation is dominated by $O(R (\prod I_k) / P)$
		\item communication is dominated by $O(R \sum I_k / P_k)$
		\item computation scales with $P$, communication scales with $P^{1/N}$, but it doesn't matter until $P$ is huge
	\end{itemize}
	\vfill
	\item Dimension trees make huge difference for higher-order tensors
	\begin{itemize}
	\scriptsize
		\item dominant computation term predicts $N/2$ speedup
		\item avoiding large Khatri-Rao products also makes a big difference
		\item DGEMM calls have more favorable shapes
	\end{itemize}
	\vfill
	\item Liavas et al.'s code has some shortcomings
	\begin{itemize}
	\scriptsize
		\item limited to 3D tensors
		\item permutes the tensor entries for each matricization 
		\item does not use dimension trees or efficient error computation
		\item this doesn't explain everything, but we don't have a time breakdown for further detail
	\end{itemize}
\end{enumerate}
